---
title: "Unit 3 For Live Session R Code"
output: html_document
editor_options: 
  chunk_output_type: console
---


# Load Required Packages
```{r}
#install.packages("tidyverse")
library(tidyverse)
#install.packages("GGally")
library(GGally)
```

# Wrangle Data: Filter to get only LM and LF and get the Position factor to have 2 levels
```{r}
FIFA = read.csv(file.choose(), header = TRUE)
FIFA2 = FIFA %>% filter(Position == "LM" | Position == "LF")
View(FIFA2)
FIFA2$Position = as.factor(as.character(FIFA2$Position))
```

# Plot matrix of only variables of interest
```{r}
FIFA2 %>% select(Acceleration, Agility, Position) %>% ggpairs(aes(color = Position))
```

# Perform T Test
```{r}
t.test(Agility~Position, data = FIFA2)
```

# Check assumptions with histograms 
```{r}
summary(FIFA2$Position)
FIFA2 %>% ggplot(aes(x = Agility, color = Position, fill = Position)) + geom_histogram(bins = 9) + facet_wrap(~Position)
FIFA2 %>% filter(Position == "LF") %>% ggplot(aes(x = Agility, color = Position, fill = Position)) + geom_histogram(bins = 9)
```

FIFA3 = mutate(FIFA, PositionGroup = if_else(Position == "GK", "GK", if_else(Position == "CB" | Position == "LCB" | Position == "LWB" | Position == "LB" | Position == "RCB" | Position == "RB" | Position == "RWB", "DEF", if_else(Position == "CDM" | Position == "RDM" | Position == "RCM" | Position == "RM" | Position == "LDM" | Position == "LCM" | Position == "LM" | Position == "RAM" | Position == "LAM" | Position == "CAM", "MID", "ATT"))))

